### 2024-04-04 docs/SOSP 2024/caching.tex
Add section on representations

\section{Representations}
\TODO{Talk about array, sparse and reorg representations here.}

@@

\begin{itemize}
  \item The scheduling language described in Section \ref{sec:schedule} allows
  for numerous possible schedules for a given model. 
  Finding a schedule with good performance is a non-trivial task.
  \item We identify a reasonable template schedule for GPUs which encompasses
  all strategies published in prior work.
  \item Describe the template schedule.
  \item Even within the variants of this template schedule, there is a significant 
  amount of variation in performance. Show the histogram. 
  \item Extensively exploring all possible parameter values for the template schedule 
  is infeasible due to the large number of parameters.
  Cite some numbers on how long it takes to explore the schedule space.
  \item List the observations we make and the heuristic we design based on these.
\end{itemize}

>>>>

\section{Representations}
\label{sec:representations}

The design of the \Treebeard{} compiler allows the implementation of different strategies for 
the in-memory representation of the model. The compiler currently has implementations for 
the three representations shown in Figure \ref{Fig:Representations}. The array and sparse 
representations are the ones described in the \Treebeard{} paper\cite{Treebeard}. The reorg 
representation is the representation used by the RAPIDs library\cite{RAPIDs}. 
The \textbf{array representation} is the simplest representation where the trees are stored in an array
in level order. The \textbf{sparse representation} stores the trees in a sparse format where only the
non-zero nodes are stored and nodes contain pointers to their children. The \textbf{reorg representation}
interleaves the array representation of each tree in the model: all root nodes are stored first, then 
the left children of all the roots and so on.
\begin{figure}[htb]
  \centering
  \includegraphics[width=\linewidth]{figures/representations.png}
  \caption{The three representations supported by \Treebeard{}.}
  \label{Fig:Representations}
\end{figure}

One of the major changes we make to the original design of \Treebeard{} \cite{Treebeard} is to
separate the implementation of representations from the rest of the compiler. This allows
us to implement representations as plugins to the compiler. We do this by defining an interface 
that each representation needs to implement. The code generator is implemented using methods 
on the representation interface thus hiding details of the actual representation from the core 
compiler. The interface abstracts the following details. 
\begin{itemize}
  \item \textbf{Buffer allocation:} The representation object exposes methods that generate 
  buffer allocations in the IR. 
  \item \textbf{Moving to child nodes:} Given the value of the predicate at a node (or tile),
  generate code to move to the appropriate child node of the current node.
  \item \textbf{Details of leaves:} The interface abstracts determining whether the current
  node is a leaf (which is needed to terminate walks) and how to get the value of the leaf
  (which is needed to get a tree's prediction).
  \item \textbf{Caching trees:} Since reading the trees into shared memory on GPU (or prefetching)
  them on CPU require knowledge of the buffer layout, the task of generating caching code for trees
  is delegated to the representation object. 
  \item \textbf{Loading thresholds and feature indices:} The representation object provides MLIR 
  rewrite patterns to lower operations that load thresholds and feature indices to LLVM IR. This
  is necessary because the details of how to load the thresholds and feature indices from the model 
  buffers is representation-specific.
\end{itemize}

In summary, the representation interface abstracts the details of how the model is stored in memory
and allows the compiler to generate code without having to explicitly know the details of the
representation. This design allows us to implement new representations without changing the core
compiler infrastructure. Implementing the representations as plugins also allows us to reuse
the implementations across different lowering pipelines. For example, the code for the array 
and sparse representations is almost fully shared between the CPU and GPU lowering pipelines.

@@

% \begin{itemize}
%   \item The scheduling language described in Section \ref{sec:schedule} allows
%   for numerous possible schedules for a given model. 
%   Finding a schedule with good performance is a non-trivial task.
%   \item We identify a reasonable template schedule for GPUs which encompasses
%   all strategies published in prior work.
%   \item Describe the template schedule.
%   \item Even within the variants of this template schedule, there is a significant 
%   amount of variation in performance. Show the histogram. 
%   \item Extensively exploring all possible parameter values for the template schedule 
%   is infeasible due to the large number of parameters.
%   Cite some numbers on how long it takes to explore the schedule space.
%   \item List the observations we make and the heuristic we design based on these.
% \end{itemize}
